---
title: Don't do PCA After Statistical Testing!
author: Robert M Flight
date: '2018-09-14'
slug: don-t-do-pca-after-statistical-testing
categories: []
tags:
  - pca
  - bioinformatics
  - R
---

```{r setup, echo = FALSE, include = FALSE}
library(broom)
library(dplyr)
library(visualizationQualityControl)
set.seed(1234)
```

## TL;DR

If you do a statistical test **before** a dimensional reduction method like
PCA, the highest source of variance is likely to be whatever you tested
statistically.

## Wait, Why??

Let me describe the situation. You've done an `-omics` level analysis on your
system of interest. You run a t-test (or ANOVA, etc) on each of the features
in your data (gene, protein, metabolite, etc). Filter down to those things that
were statistically significant, and then finally, you decide to look at the data
using a dimensionality reduction method such as *principal components analysis*
(PCA) so you can **see** what is going on.

I have seen this published at least once (in a Diabetes metabolomics paper,
if anyone knows it, please send it to me so I can link it), and have seen
collaborators do this after coaching from others in non-statistical departments.

## The Problem

The problem is that PCA is just looking at either feature-feature covariances
or sample-sample covariances. If you have trimmed the data to those things that
have statistically significant differences, then you have completely modified
the covariances, and PCA is likely to pick up on that.

## An Example

Let's actually do an example where there are no differences initially, and
then see if we can introduce an artificial difference.

### Random Data

We start with completely random data, 10000 features, and 100 samples.

```{r setup_data}
n_feat = 10000
n_sample = 100
random_data = matrix(rnorm(n_feat * n_sample), nrow = n_feat, ncol = n_sample)
```

Now we will do a t-test on each row, taking the first 50 samples as class 1
and the other 50 samples as class 2.

```{r do_t_test}
t_test_res = purrr::map_df(seq(1, nrow(random_data)), function(in_row){
  tidy(t.test(random_data[in_row, 1:50], random_data[in_row, 51:100]))
})
```

How many are significant at a p-value of 0.05?

```{r sig_p_values}
filter(t_test_res, p.value <= 0.05) %>% dim()
```

Obviously, these are false positives, but they are enough for us to illustrate
the problem.

First, lets do PCA on the whole data set of 10000 features.

```{r all_pca, message = FALSE, warning = FALSE}
sample_classes = data.frame(class = c(rep("A", 50), rep("B", 50)))

all_pca = prcomp(t(random_data), center = TRUE, scale. = FALSE)
visqc_pca(all_pca, groups = sample_classes$class)
```

Obviously, there is no difference in the groups, and the % explained variance
is very low.

Second, lets do it on just those things that were significant:

```{r sig_pca, message = FALSE, warning = FALSE}
sig_pca = prcomp(t(random_data[which(t_test_res$p.value <= 0.05), ]), center = TRUE,
                 scale. = FALSE)

visqc_pca(sig_pca, groups = sample_classes$class)
```

And look at that! We have separation of the two groups! But ...., this is
completely random data, that didn't have any separation, **until we did the
statistical test**!

## Take Away

Be careful of the order in which you do things. If you want to do dimensionality
reduction to look for issues with the samples, then do that **before** any statistical
testing on the individual features.
